---
title: "7.3 Inference and Linear Regression (R3)"
author: ""
output: html_notebook
---

The slope and vertical intercept, $\beta_1, \beta_0$ we computed in Section 7.2 are determined by a sample of (pairs of) random variables, and as such, are merely point estimates of these parameters. As usual, there is some chance we chose non-representative data from the sample.
In this section, we use our sample data to make inferences on the true parameters.

### Exploration 7.3.1. Estimating Regression. 

Let X and Y be variables where $Y=0.8X+5+\epsilon$ is a distributed by a normal variable with mean 0 and standard deviation 3.

(a)
What line probably best fits the relationship between X and Y?

(b)
Run the following code to generate n=10 random X and Y values, plot them, find a regression line and the line in (a): 
```{r}
n=10
X=runif(n, 0, 10)
Y=0.8*X+5+rnorm(n,0,3)
mod=lm(Y~X)
plot(X, Y, pch=19)
abline(mod)
abline(5, 0.8, col="blue", lty=2)
```

Run it a few times, what do you notice?

(c)
Adjust n=100 and run it again, now what do you notice?

## Subsection 7.3.1 Hypothesis Testing 

### Remark 7.3.1. 

When it comes to linear regression, we focus on the following pair of hypothesis. :

*  $H_0: \beta_1=0$ or the slope is zero. So changes in the explanatory variable do not result in average change in response variable.

*  $H_0: \beta_1\neq 0$ or the slope is not zero. So changes in the explanatory variable do result in average change in response variable.

This is a numerical hypothesis test similar to what we have done before. However we're not just given a list of slopes to find a sample standard deviation. The computation here is tedious, so we utilize technology to perform them. If one wished to compute the standard error for $\beta_1$ by hand, we do so via:

$SE_1=\sqrt{\frac{1}{n-2}\cdot \frac{\sum (y_i-\hat{y}_i)^2}{\sum(x_i-\bar{x})^2}}$

where $\hat{y}_i=\beta_1\cdot x_i+\beta_0$, or the value of $y$ predicted by the linear model.

### Exploration 7.3.2. Hypothesis Testing: Slope - Possums. 

Run the following code to download `possum.csv` as seen in Exploration 7.1.1, Activity 7.1.7, and Exploration 7.2.1, and create a linear model for `head_l~skull_w` and summarize it. 
```{r}
possum = read.csv("https://github.com/TienChih/tbil-stats/raw/main/data/possum.csv")
possummod=lm(head_l~skull_w, data=possum)
summary(possummod)
```

We're focused on the `Coefficients`, in particular the second row.
The first row starting with (`Intercept`) gives the point estimate, standard error, test statistic and $p$-value for the alternative hypothesis $H_A:\beta_0\neq 0$.

The second row starting with `skull_w` gives the point estimate, standard error, test statistic and $p$-value for the alternative hypothesis $H_A:\beta_1\neq 1$. This is the row about slopes.

(a)
The first entry of this row gives us the point estimate for the slope. How does this compare to slope found by running: 
```{r}
possummod
beta_1 = possummod$coefficients[2]
```

We call this value $\beta_1$.

(b)
The second entry of this row gives us the standard error for the slope. Call this value $SE_1$. Find the $t$-value for $\beta_1$ by computing
$t_1=\frac{\beta_1-0}{SE_1}.$
```{r}
SE_1 = sqrt(diag(vcov(possummod)))[2]
t_1 = (possummod$coefficients[2] - 0)/SE_1
```

How does this value compare to the third entry?

(c)
According to the summary, how many degrees of freedom are there?
```{r}
df = summary(possummod)$df[2]
```

(d)
Compute $P(t>t_1)$ on the standard $t$-distribution with the appropriate degrees of freedom. Then double it to obtain the $p$-value for the alternative hypothesis $H_A: \beta_1\neq 0$. 
```{r}
pt(t_1, df, lower.tail = FALSE)*2
```


How does this value compare to the 4th entry?

(e)
Do we accept or reject the null hypothesis that $H_0: \beta_1=0$?

### Remark 7.3.2. 

The fourth entry of the second Coefficients: row gives the probability that, if the slope was 0, that we would obtain a slope as steep or steeper.

The fourth entry of the first Coefficients: row gives the probability that, if the intercept was 0, that we would obtain an as or more extreme.

## Subsection 7.3.2 Confidence Intervals 

### Remark 7.3.3. 

For $\beta_i$, we compute the C % confidence interval as follows:
$[\text{point estimate for }\beta_1 - SE_it^*, \text{point estimate for }\beta_1 + SE_it^*]$
where $t^*$ is the $t$-value such that $P(-t^*\lt t\lt t^*)=C\%$ for the standard $t$-variable with the appropriate degrees of freedom.

### Exploration 7.3.3. Confidence: Intervals:$\beta_i$ - Possums. 

We continue from Exploration 7.3.2.

(a)
Find a $t^*$ so that $P(-t^*\lt t\lt t^*)=C\%$ for the appropriate degrees of freedom.
```{r}
tstar = qt(0.975, df)
```

(b)
Use the point estimate and standard error for $\beta_1$, $t^*$ and Remark 7.3.3 to compute a 95% confidence interval for $\beta_1$.
```{r}
beta_1 - tstar*SE_1
beta_1 + tstar*SE_1
```

(c)
Explain what this confidence interval means within the context of the problem.

## Subsection 7.3.3 Putting it together 

### Activity 7.3.4. Inference for SP500 companies. 

Run the following code to download `sp500.csv` a data set containing information on a sample of 50 fortune 500 companies and show it's variable names. 
```{r}
sp500 = read.csv("https://github.com/TienChih/tbil-stats/raw/main/data/sp500.csv")

names(sp500)
```

Click here to learn more about this data set: https://www.openintro.org/data/index.php?data=sp500.

(a)
Run the following to create and summarize a linear model with debt the debt in millions of dollars as the explanatory variable and profit_margin the percent of earnings that is profit as the response variable: 
```{r}
sp500mod=lm(profit_margin~debt, data=sp500)
summary(sp500mod)
```

(b)
Run the following to plot profit_margin of these companies against the debt and draw a regression line: 
```{r}
plot(sp500$debt, sp500$profit_margin)
abline(sp500mod, col="red")
```

(c)
State what the slope $\beta_1$ means in the context of this problem.

(d)
Interpret the $p$-value for the alternative hypothesis $H_A:\beta_1\neq 0$ in the context of this problem.

(e)
Do we reject the null hypothesis that $H_0:\beta_1=0$? What does that say about the relationship between profit margin and debt?

(f)
Find a $t^*$ so that $P(-t^*\lt t\lt t^*)=C\%$ for the appropriate degrees of freedom.
```{r}
tstar = qt(1-(1-0.95)/2, summary(sp500mod)$df[2],lower.tail = TRUE)
tstar
```

(g)
Use the point estimate and standard error for $\beta_1$, $t^*$ and Remark 7.3.3 to compute a 95% confidence interval for $\beta_1$.
```{r}
SE_1 = sqrt(diag(vcov(sp500mod)))[2]
beta_1 = sp500mod$coefficients[2]

beta_1 - tstar*SE_1
beta_1 + tstar*SE_1
```


(h)
Explain what this confidence interval means within the context of the problem.


### Quiz 6 Activity

Run the following code to enter class data on Chapter 6 Homework totals and Chapter 6 Quiz scores
```{r}
Homework = c( 0, 0, 5, 7, 7.38, 40.03, 52.34, 52.84, 58.28, 58.95, 73.29, 73.89, 74.89, 74.94, 80.55, 81.84, 84.39, 84.89, 91.5)
Quiz = c(8.33, 3.33, 0.21, 6.88, 2.71, 16.04, 6.46, 2.71, 8.33, 17.29, 7.5, 6.67, 9.58, 10.83, 19.79, 17.92, 24.38, 11.88, 23.96)

```

(a)  Create a linear model using homework totals to explain quiz scores.
```{r}
quizmod = lm(Quiz ~ Homework)
```

(b) What can you tell from the summary of the model? 

(c) What percentage of the quiz score is explained by the homework score? 

(d) State what the slope $\beta_1$ means in the context of this problem.

(e) Interpret the $p$-value for the alternative hypothesis $H_A:\beta_1\neq 0$ in the context of this problem.

(f) Do we reject the null hypothesis that $H_0:\beta_1=0$? What does that say about the relationship between profit margin and debt?

(g) Find a $t^*$ so that $P(-t^*\lt t\lt t^*)=C\%$ for the appropriate degrees of freedom.
```{r}
tstar = qt(1-(1-0.95)/2, summary(quizmod)$df[2],lower.tail = TRUE)
tstar
```

(h) Use the point estimate and standard error for $\beta_1$, $t^*$ and Remark 7.3.3 to compute a 95% confidence interval for $\beta_1$.
```{r}
SE_1 = sqrt(diag(vcov(quizmod)))[2]
beta_1 = quizmod$coefficients[2]

beta_1 - tstar*SE_1
beta_1 + tstar*SE_1
```


(i) Explain what this confidence interval means within the context of the problem.

(j) Are there any lessons learned from this data?


### Activity 7.3.5. Inference for Nutrition and Starbucks. 
Run the following code to download starbucks.csv a data set containing information about 77 Starbucks menu items their nutritional value and show it's variable names. 
```{r}
starbucks = read.csv("https://github.com/TienChih/tbil-stats/raw/main/data/starbucks.csv")

names(starbucks)
```

Click here to learn more about this data set: https://www.openintro.org/data/index.php?data=starbucks.

(a)
Run the following to create and summarize a linear model with protein the protein content of an item in g as the explanatory variable and calories the calories of each item measure in, well calories, as the response variable: 
```{r}
starbucksmod=lm(calories~protein, data=starbucks)
summary(starbucksmod)
```

(b)
Run the following to plot calories of these items against the protein and draw a regression line: 
```{r}
plot(starbucks$protein, starbucks$calories)
abline(starbucksmod, col="red")
```

(c)
State what the slope $\beta_1$ means in the context of this problem.
```{r}
beta_1 = starbucksmod$coefficients[2]
beta_1
```

(d)
Interpret the $p$-value for the alternative hypothesis $H_A:\beta_1\neq 0$ in the context of this problem.
```{r}

```

(e)
Do we reject the null hypothesis that $H_0:\beta_1=0$? What does that say about the relationship between calories and protein?

(f)
Find a $t^*$ so that $P(-t^*\lt t\lt t^*)=C\%$ for the appropriate degrees of freedom.
```{r}
tstar = qt(1-(1-0.95)/2, summary(FIXME)$df[2],lower.tail = TRUE)
tstar
```

(g)
Use the point estimate and standard error for $\beta_1$, $t^*$ and Remark 7.3.3 to compute a 95% confidence interval for $\beta_1$.
```{r}
SE_1 = sqrt(diag(vcov(FIXME)))[2]

beta_1 - tstar*SE_1
beta_1 + tstar*SE_1

```


(h)
Explain what this confidence interval means within the context of the problem.

(i)
Repeat this for any other pair of numerical variables.
