---
title: "Short Section 5-2"
author: "Mariah Birgen"
output: html_notebook
---
### Initial Information
Input the information from your sample:
```{r}
x_1 = FIXME
n_1 = FIXME
x_2 = FIXME
n_2 = FIXME
    
```

### Calculating Confidence Intervals:
If you sample n_1 items and have x_1 positives and you sample n_2 other items and have x_2 positives and you want a C-confidence interval
```{r}
phat_1=x_1/n_1
phat_2=x_2/n_2

d=phat_1-phat_2

SE_D=sqrt(phat_1*(1-phat_1)/n_1 + phat_2*(1-phat_2)/n_2)
z_star = qnorm(1-C/2,0,1)
LB=d-z_star*SE_D
UB=d+z_star*SE_D
print(paste0("Lower Bound: ", LB , ", Upper Bound:", UB))
```

### Calculating Hypothesis Tests

Make sure you have input your information above and have decided what type of hypothesis test you are calculating.

#### Pooled Samples
If your null hypothesis is that both populations have the same proportion, we need to calculate the pooled sample proportion.
```{r}
p_pool = (x_1+x_2)/(n_1+n_1)
```

$\mu_D$ is the mean of your null hypothesis.
```{r}
mu_D = FIXME
SE_D = sqrt(p_pool*(1-p_pool)/n_1+p_pool*(1-p_pool)/n_2)
observed_diff = x_1/n_1-x_2/n_2
observed_diff
mu_D
SE_D
```

Next, calculate $\hat z$
```{r}
z_hatp = (observed_diff-mu_D)/SE_D
print(z_hatp)
```

Last, calculate your p-value. 
If your alternative hypothesis is one-sided:
```{r}
pnorm(abs(z_hatp),0,1,lower.tail = FALSE)
```

If your alternative hypothesis is two-sided:
```{r}
2*pnorm(abs(z_hatp),0,1,lower.tail = FALSE)
```

Run the following code to n_1 of your first population and n_2 of your second population using the pooled proportion. It does this `trials=1000` times and plots a histogram of the differences.
```{r}
trials=1000
num_men=n_1
num_women=n_2
m_raise=x_1
w_raise=x_2

raise_rate=(m_raise+w_raise)/(num_men+num_women)
diffraise=rep(NA, trials)

for(i in 1:trials){
    menraise=sample(c("RAISE", "NO"), num_men, replace=TRUE, prob=c(raise_rate, 1-raise_rate))
    womenraise=sample(c("RAISE", "NO"), num_women, replace=TRUE, prob=c(raise_rate, 1-raise_rate))
    diffraise[i]=length(which(menraise=="RAISE"))/num_men-length(which(womenraise=="RAISE"))/num_women
}
hist(diffraise, breaks=25, prob=TRUE)
```

The following code to plot a normal curve with mean `mu_D` and standard deviation `SE_D`, and plot a line for the observed difference `observed_diff=`
```{r}

hist(diffraise, breaks=25, prob=TRUE)
curve(dnorm(x, mean=mu_D, sd=SE_D), col="darkblue", lwd=2, add=TRUE, yaxt="n")
abline(v = observed_diff, col="red", lwd=3, lty=2)

```

Run the following code to see what proportion of these differences are greater than or equal to the observed_diff. 
```{r}
length(which(diffraise>=observed_diff))/trials
```

### Short Code
The possible alternatives are "two.sided", "greater" and "less".

```{r}
prop.test(c(x_1, x_2), c(n_1, n_2), 
        p=NULL, alternative="greater", correct=FALSE)
```